---
title: "Is there an optimal time and place to fill up your vehicle in Victoria?"
subtitle: "Assignment 4 ETC5512"
author: "Pei Sin Ang"
format: html
editor: visual
execute: 
  message: false
  warning: false
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

**Topic:**

Is there an optimal time and place to fill up your vehicle in Victoria?

- Does an area's fuel price reflect its geographic location, or is it driven by the brands available nearby?

- Do socioeconomically disadvantaged areas pay more or less for fuel compared to wealthier areas?

- What is the optimal day of the week to refuel?

While browsing the [ABC's national fuel price tracker](https://www.abc.net.au/news/2026-04-01/petrol-diesel-price-tracker/106513484) i noticed that Victoria's data was absent, due to the way the state releases price information. As someone personally affected by the fluctuating fuel costs, I wanted to fill that gap and investigate whether there are smart times and places to refuel in Victoria, using the Servo Saver dataset.

**Dataset:**

1.  [Servo Saver Public API](https://service.vic.gov.au/find-services/transport-and-driving/servo-saver/help-centre/servo-saver-public-api#h2-11)

    - Source: Service Victoria

    - Content: Fuel price data for all registered service stations across Victoria

    - Last Updated: 24-hour delay released

    - Data Type: Census

    - Licence: Creative Commons Attribution 4.0 International licence

    - Limitation: Data refresh every 24-hour, there is no historical data. User will have to pull the data from the API key periodically. Coverage depends on retailer compliance.

    - Privacy/Ethics: Not personally identifiable information. Commercial data only.
2.  [ABS Digital Boundary Files](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs/edition-3-july-2021-june-2026/access-and-downloads/digital-boundary-files)

    - Source: Australian Bureau of Statistics (ABS)

    - Content: Spatial boundary shapefiles used to map service station locations to statistical areas for geographic price comparison.
    
    - Last updated: 2021

    - Data Type: GeoPackage/Shapefile

    - Licence: Creative Commons Attribution 4.0 International licence

    - Limitation: Boundaries released in 2021 and may not reflect recent urban developments.
    
    - Privacy/Ethics: No privacy concerns, geographic reference data only.

3. [Socio-Economic Indexes for Areas (SEIFA), Australia 2021](https://www.abs.gov.au/statistics/people/people-and-communities/socio-economic-indexes-areas-seifa-australia/latest-release)

    - Source: Australian Bureau of Statistics (ABS)
    
    - Content: Uses 2021 Census data to summarise the socio-economic characteristics of an area (such as income, education, occupation, housing and family structure)
    
    - Last updated: 27 April 2023 (refences 2021 Census data)
    
    - Data Type: Derived census-based aggregate
    
    - Licence: Creative Commons Attribution 4.0 International licence
    
    - Limitations: Based on 2021 Census data and may not reflect current socioeconomic conditions.
    
    - Privacy/Ethics: Aggregated area-level data, no reidentification concern.

**Servo Saver API:**

1. Obtain API access and authorisation from [Servo Saver Public API webpage](https://service.vic.gov.au/find-services/transport-and-driving/servo-saver/help-centre/servo-saver-public-api/apply-for-servo-saver-public-api)
    - Access will take at least 24 hours, and the API key will be sent to applicants via email upon approval.

2. Pull data
```{r}
#| eval: false

library(httr2)

# Request call 

req <- request("https://api.fuel.service.vic.gov.au/open-data/v1/fuel/prices") |> 
  req_headers(
    "User-Agent" = "MyApp/1.0",
    "x-consumer-ID" = Sys.getenv("SERVO_API_KEY"),
    "x-transactionid" = "6ff20493-4de0-4735-a036-e1e8dd41b33c"
  ) |> 
  req_user_agent("R httr2 servo-saver-client")

# Getting response 

resp <- req_perform(req) |> 
  resp_body_json(simplifyVector = TRUE) 

# Save raw data as a tibble
tibble::as_tibble(resp)
```

3. Unnest raw data that has dataframes within rows (repeat for every day the data is extracted) - Data extracted at 11.50pm AEST everyday from 1 June to 5 June (Monday to Friday)
```{r}
#| eval: false

fuel_df_raw_1Jun <- resp$fuelPriceDetails |>
  as_tibble() |>
  unnest_wider(fuelStation) |>
  rename(station_updatedAt = updatedAt) |>
  unnest(fuelPrices) |> 
  unnest(location)

saveRDS(fuel_df_raw_1Jun, "data/fuel_df_raw_2026-06-01.rds")
```

4. Combine all 5 days of data into one
```{r}
#| eval: false

library(tidyverse)

fuel_raw_1Jun <- readRDS("data/fuel_df_raw_2026-06-01.rds")
fuel_raw_2Jun <- readRDS("data/fuel_df_raw_2026-06-02.rds")
fuel_raw_3Jun <- readRDS("data/fuel_df_raw_2026-06-03.rds")
fuel_raw_4Jun <- readRDS("data/fuel_df_raw_2026-06-04.rds")
fuel_raw_5Jun <- readRDS("data/fuel_df_raw_2026-06-05.rds")

fuel_raw_combined <- bind_rows(fuel_raw_1Jun, 
                               fuel_raw_2Jun,
                               fuel_raw_3Jun,
                               fuel_raw_4Jun,
                               fuel_raw_5Jun
                               )
```

5. Process data and save
```{r}
#| eval: false

fuel_df_clean <-fuel_raw_combined |> 
  select(-openingHours, -id) |>
  # Change timezone to Melbourne Time
  mutate(
    station_updatedAt_dt = ymd_hms(station_updatedAt, tz = "UTC") |>
      with_tz("Australia/Melbourne"), 
    station_date = as.Date(station_updatedAt_dt, tz = "Australia/Melbourne"),
    station_time = format(station_updatedAt_dt, "%H:%M:%S")
  ) |> 
  select(-station_updatedAt_dt, -station_updatedAt) |> 
  # Only assessing U91 and Diesel 
  filter(fuelType %in% c("U91", "DSL")) 

# Save processed data 
saveRDS(fuel_df_clean, "data/fuel_df_clean.rds")
```

**ABS Digital Boundary File and SEIFA**

1. Navigate to [ABS Digital Boundary Files](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs/edition-3-july-2021-june-2026/access-and-downloads/digital-boundary-files)

2. Download the Statistical Area Level 1 (SA1) - 2021 - Shapefile

3. Navigate to [Socio-Economic Indexes for Areas (SEIFA), Australia 2021](https://www.abs.gov.au/statistics/people/people-and-communities/socio-economic-indexes-areas-seifa-australia/latest-release)

4. Download the Statistical Area Level 1, Indexes, SEIFA 2021.xlsx

5. Process and save SA1 shapefile
```{r}
#| eval: false
# Read raw shapefiles
sa1_raw <- st_read("data/SA1_2021_AUST_SHP_GDA2020/SA1_2021_AUST_GDA2020.shp")

# Filter to Victoria only
sa1_vic <- sa1_raw |> 
  filter(STE_CODE21 == "2") |> 
  select(SA1_CODE21, geometry)

saveRDS(sa1_vic, "data/sa1_vic.rds")
```

6. Read, process, join and save the SEIFA SA1 data
```{r}
#| eval: false

library(readxl)
seifa_raw <- read_excel("data/Statistical Area Level 1, Indexes, SEIFA 2021.xlsx",
                        sheet = "Table 1",
                        skip = 5)
# Rename header
seifa_clean <- seifa_raw |>
  rename(
    SA1_CODE21    = 1,
    IRSD_score    = 2,
    IRSD_decile   = 3,
    IRSAD_score   = 4,
    IRSAD_decile  = 5,
    IER_score     = 6,
    IER_decile    = 7,
    IEO_score     = 8,
    IEO_decile    = 9
  ) |>
  select(SA1_CODE21, IRSD_score, IRSD_decile, IRSAD_score, IRSAD_decile,
         IER_score, IER_decile, IEO_score, IEO_decile) |>
  mutate(SA1_CODE21 = as.character(SA1_CODE21)) |>
  filter(startsWith(SA1_CODE21, "2")) #filter only Victoria SA1

# Save processed SEIFA data
saveRDS(seifa_clean, "data/seifa_clean.rds")

# Sanity check before joining 
sum(sa1_vic$SA1_CODE21 %in% seifa_clean$SA1_CODE21) 

# Join data
sa1_seifa <- sa1_vic |>
  left_join(seifa_clean, by = "SA1_CODE21")

# Save joined data 
saveRDS(sa1_seifa, "data/sa1_seifa.rds")
```


<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### What's in this section

Here is where you should write your blogpost! (Task 2)

### Blogpost Structure

There is no strict structure for you to follow. However, here is a skeleton structure you may find helpful.

1.  Title (is set at the top of this document)
2.  Motivation\
3.  Data\
4.  Analysis\
5.  Conclusions\
6.  References

**What is the optimal day of the week to refuel?**

```{r}
library(lubridate)

fuel_df_clean <- readRDS("data/fuel_df_clean.rds")

fuel_df_clean |>
  mutate(transaction_day = factor(wday(station_date, label = TRUE, abbr = FALSE))) |>
  ggplot(aes(x = transaction_day, y = price, fill = transaction_day)) +
  geom_boxplot() +
  facet_wrap(~ fuelType) +
  scale_fill_brewer(palette = "Set2") +
  theme_bw() +
  labs(x = "Day of Week", y = "Price", fill = "Day of Week")
```

**Does an area's fuel price reflect its geographic location, or is it driven by the brands available nearby?**

```{r}
sa1_seifa <- readRDS("data/sa1_seifa.rds")

fuel_sf <- fuel_df_clean |>
  distinct(latitude, longitude, .keep_all = TRUE) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(crs = st_crs(sa1_seifa))

ggplot() +
  geom_sf(data = sa1_seifa, colour = "grey70", fill = "grey95", linewidth = 0.1) +
  geom_sf(data = fuel_sf, colour = "steelblue", size = 0.5, alpha = 0.6) +
  coord_sf(xlim = c(144.5, 145.5), ylim = c(-38.2, -37.6)) +
  theme_bw() +
  labs(title = "Fuel Stations across Victoria SA1 Areas")

ggplot() +
  geom_sf(data = sa1_seifa, aes(fill = as.numeric(IRSD_decile)), 
          colour = NA, linewidth = 0) +
  scale_fill_viridis_c() +
  coord_sf(xlim = c(144.5, 145.5), ylim = c(-38.2, -37.6))

```

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.
:::
