---
title: "What is your blogpost title? Add it here??"
subtitle: "Assignment 4 ETC5512"
author: "Pei Sin Ang"
format: html
editor: visual
execute: 
  message: false
  warning: false
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

**Topic:**

Is there an optimal time and place to fill up your vehicle in Victoria?

- Does an area's fuel price reflect its geographic location, or is it driven by the brands available nearby?

- Do socioeconomically disadvantaged areas pay more or less for fuel compared to wealthier areas?

- What is the optimal day of the week to refuel?

While browsing the [ABC's national fuel price tracker](https://www.abc.net.au/news/2026-04-01/petrol-diesel-price-tracker/106513484) i noticed that Victoria's data was absent, due to the way the state releases price information. As someone personally affected by the fluctuating fuel costs, I wanted to fill that gap and investigate whether there are smart times and places to refuel in Victoria, using the Servo Saver dataset.

**Dataset:**

1.  [Servo Saver Public API](https://service.vic.gov.au/find-services/transport-and-driving/servo-saver/help-centre/servo-saver-public-api#h2-11)

    - Source: Service Victoria

    - Content: Fuel price data for all registered service stations across Victoria

    - Last Updated: 24-hour delay released

    - Data Type: Census

    - Licence: Creative Commons Attribution 4.0 International licence

    - Limitation: Data refresh every 24-hour, there is no historical data. User will have to pull the data from the API key periodically. Coverage depends on retailer compliance.

    - Privacy/Ethics: Not personally identifiable information. Commercial data only.
2.  [ABS Digital Boundary Files](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs/edition-3-july-2021-june-2026/access-and-downloads/digital-boundary-files)

    - Source: Australian Bureau of Statistics (ABS)

    - Content: Spatial boundary shapefiles used to map service station locations to statistical areas for geographic price comparison.
    
    - Last updated: 2021

    - Data Type: GeoPackage/Shapefile

    - Licence: Creative Commons Attribution 4.0 International licence

    - Limitation: Boundaries released in 2021 and may not reflect recent urban developments.
    
    - Privacy/Ethics: No privacy concerns, geographic reference data only.

3. [Socio-Economic Indexes for Areas (SEIFA), Australia 2021](https://www.abs.gov.au/statistics/people/people-and-communities/socio-economic-indexes-areas-seifa-australia/latest-release)

    - Source: Australian Bureau of Statistics (ABS)
    
    - Content: Uses 2021 Census data to summarise the socio-economic characteristics of an area (such as income, education, occupation, housing and family structure)
    
    - Last updated: 27 April 2023 (refences 2021 Census data)
    
    - Data Type: Derived census-based aggregate
    
    - Licence: Creative Commons Attribution 4.0 International licence
    
    - Limitations: Based on 2021 Census data and may not reflect current socioeconomic conditions.
    
    - Privacy/Ethics: Aggregated area-level data, no reidentification concern.

**Servo Saver API:**

1. Obtain API access and authorisation from [Servo Saver Public API webpage](https://service.vic.gov.au/find-services/transport-and-driving/servo-saver/help-centre/servo-saver-public-api/apply-for-servo-saver-public-api)
    - Access will take at least 24 hours, and the API key will be sent to applicants via email upon approval.

2. Pull data
```{r}
#| eval: false

library(httr2)

# Request call 

req <- request("https://api.fuel.service.vic.gov.au/open-data/v1/fuel/prices") |> 
  req_headers(
    "User-Agent" = "MyApp/1.0",
    "x-consumer-ID" = Sys.getenv("SERVO_API_KEY"),
    "x-transactionid" = "6ff20493-4de0-4735-a036-e1e8dd41b33c"
  ) |> 
  req_user_agent("R httr2 servo-saver-client")

# Getting response 

resp <- req_perform(req) |> 
  resp_body_json(simplifyVector = TRUE) 

# Save raw data as a tibble
tibble::as_tibble(resp)
```

3. Unnest raw data that has dataframes within rows (repeat for every day the data is extracted) - Data extracted at 11.50pm AEST everyday from 1 June to 5 June (Monday to Friday)
```{r}
#| eval: false

fuel_df_raw_1Jun <- resp$fuelPriceDetails |>
  as_tibble() |>
  unnest_wider(fuelStation) |>
  rename(station_updatedAt = updatedAt) |>
  unnest(fuelPrices) |> 
  unnest(location)

saveRDS(fuel_df_raw_1Jun, "data/fuel_df_raw_2026-06-01.rds")
```

4. Combine all 5 days of data into one
```{r}
#| eval: false

library(tidyverse)

fuel_raw_1Jun <- readRDS("data/fuel_df_raw_2026-06-01.rds")
fuel_raw_2Jun <- readRDS("data/fuel_df_raw_2026-06-02.rds")
fuel_raw_3Jun <- readRDS("data/fuel_df_raw_2026-06-03.rds")
fuel_raw_4Jun <- readRDS("data/fuel_df_raw_2026-06-04.rds")
fuel_raw_5Jun <- readRDS("data/fuel_df_raw_2026-06-05.rds")

fuel_raw_combined <- bind_rows(fuel_raw_1Jun, 
                               fuel_raw_2Jun,
                               fuel_raw_3Jun,
                               fuel_raw_4Jun,
                               fuel_raw_5Jun
                               )
```

5. Process data and save
```{r}
#| eval: false

fuel_df_clean <-fuel_raw_combined |>
  select(-openingHours, -id) |>
  # Change timezone to Melbourne Time
  mutate(
    station_updatedAt_dt = ymd_hms(station_updatedAt, tz = "UTC") |>
      with_tz("Australia/Melbourne"),
    station_date = as.Date(station_updatedAt_dt, tz = "Australia/Melbourne"),
    station_time = format(station_updatedAt_dt, "%H:%M:%S")
  ) |>
  select(-station_updatedAt_dt, -station_updatedAt) |>
  filter(!is.na(latitude), !is.na(longitude), !is.na(price),
         fuelType %in% c("U91", "DSL"))

# Save processed data 
saveRDS(fuel_df_clean, "data/fuel_df_clean.rds")
```

**ABS Digital Boundary File and SEIFA**

1. Navigate to [ABS Digital Boundary Files](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs/edition-3-july-2021-june-2026/access-and-downloads/digital-boundary-files)

2. Download the Suburb and Localities (SAL) - 2021 - Shapefile

3. Navigate to [Socio-Economic Indexes for Areas (SEIFA), Australia 2021](https://www.abs.gov.au/statistics/people/people-and-communities/socio-economic-indexes-areas-seifa-australia/latest-release)

4. Download the Suburb and Localities, Indexes, SEIFA 2021.xlsx

5. Process and save SAL shapefile
```{r}
#| eval: false
library(sf)

# Read raw shapefiles
sal_raw <- st_read("data/SAL_2021_AUST_GDA2020_SHP/SAL_2021_AUST_GDA2020.shp")

# Filter to Victoria only
sal_vic <- sal_raw |>
  filter(STE_CODE21 == "2") |>
  select(SAL_CODE21, AREASQKM21, geometry)

saveRDS(sal_vic, "data/sal_vic.rds")
```

6. Read, process, join and save the SEIFA SAL data
```{r}
#| eval: false
library(readxl)

seifa_sal_raw <- read_excel("data/Suburbs and Localities, Indexes, SEIFA 2021.xlsx",
                            sheet = "Table 1",
                            skip = 5)

# Rename headers
seifa_sal_clean <- seifa_sal_raw |>
  rename(
    SAL_CODE21   = 1,
    SAL_name     = 2,
    IRSD_score   = 3,
    IRSD_decile  = 4,
    IRSAD_score  = 5,
    IRSAD_decile = 6,
    IER_score    = 7,
    IER_decile   = 8,
    IEO_score    = 9,
    IEO_decile   = 10
  ) |>
  select(SAL_CODE21, SAL_name, IRSD_score, IRSD_decile, IRSAD_score, IRSAD_decile,
         IER_score, IER_decile, IEO_score, IEO_decile) |>
  mutate(SAL_CODE21 = as.character(SAL_CODE21)) |>
  filter(startsWith(SAL_CODE21, "2")) # filter only Victoria SAL

# Save processed SEIFA SAL data
saveRDS(seifa_sal_clean, "data/seifa_sal_clean.rds")

# Sanity check before joining
sum(sal_vic$SAL_CODE21 %in% seifa_sal_clean$SAL_CODE21)

# Join data
sal_seifa <- sal_vic |>
  left_join(seifa_sal_clean, by = "SAL_CODE21")

# Save joined data
saveRDS(sal_seifa, "data/sal_seifa.rds")
```

7. Process and save SA4 shapefile
```{r}
#| eval: false

sa4_raw <- st_read("data/SA4_2021_AUST_SHP_GDA2020/SA4_2021_AUST_GDA2020.shp")

sa4_vic <- sa4_raw |>
  filter(STE_CODE21 == "2") |>
  select(SA4_CODE21, SA4_NAME21, geometry)

saveRDS(sa4_vic, "data/sa4_vic.rds")
```

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### What's in this section

Here is where you should write your blogpost! (Task 2)

### Blogpost Structure

There is no strict structure for you to follow. However, here is a skeleton structure you may find helpful.

1.  Title (is set at the top of this document)
2.  Motivation\
3.  Data\
4.  Analysis\
5.  Conclusions\
6.  References

**What is the optimal day of the week to refuel?**

Write up explanantion for optimal day of week to refuel
```{r}
#| label: fig-price-by-day
#| fig-cap: "Distribution of fuel prices by day of week for U91 and diesel"

library(tidyverse)
library(lubridate)

fuel_df_clean <- readRDS("data/fuel_df_clean.rds")

fuel_df_clean |>
  mutate(transaction_day = factor(wday(station_date, label = TRUE, abbr = FALSE))) |>
  ggplot(aes(x = transaction_day, y = price, fill = transaction_day)) +
  geom_boxplot() +
  facet_wrap(~ fuelType) +
  scale_fill_brewer(palette = "Set2") +
  theme_bw() +
  labs(x = "Day of Week", y = "Price (c/L)", fill = "Day of Week")
```

**Does an area's fuel price reflect its geographic location?**

```{r}
# Setup spatial data

library(sf)

sal_seifa <- readRDS("data/sal_seifa.rds")
sa4_vic <- readRDS("data/sa4_vic.rds")

# Spatial join SAL to SA4 to get SA4 name for each suburb
sal_with_sa4 <- sal_seifa |>
  st_join(sa4_vic[c("SA4_CODE21", "SA4_NAME21")], largest = TRUE)

# Convert all fuel observations to sf, retaining fuelType and price
fuel_sf <- fuel_df_clean |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(crs = st_crs(sal_seifa))

# Spatially join stations to SAL, then calculate median price per SAL per fuel type
sal_median_price <- sal_seifa |>
  st_join(fuel_sf) |>
  st_drop_geometry() |>
  group_by(SAL_CODE21, fuelType) |>
  summarise(median_price = median(price, na.rm = TRUE), .groups = "drop")

# Rejoin median prices back to SAL geometry
sal_price_map <- sal_seifa |>
  left_join(sal_median_price, by = "SAL_CODE21")

# Join SA4 region name into sal_price_map
sal_price_map_sa4 <- sal_price_map |>
  left_join(
    sal_with_sa4 |> st_drop_geometry() |> select(SAL_CODE21, SA4_NAME21) |> distinct(),
    by = "SAL_CODE21"
  )

# 50km radius buffer around Melbourne CBD (centroid of ABS Melbourne SAL)
melb_cbd <- sal_seifa |>
  filter(SAL_name == "Melbourne") |>
  st_centroid() |>
  st_transform(crs = 7855) # GDA2020 / MGA Zone 55 for accurate metre-based distance

melb_50km <- st_buffer(melb_cbd, dist = 50000) |>
  st_transform(crs = st_crs(sal_price_map))

melb_bbox <- st_bbox(melb_50km)
```

```{r}
#| label: fig-u91-map
#| fig-cap: "Median U91 petrol price by suburb/locality within 50km of Melbourne CBD"

ggplot() +
  geom_sf(data = sal_seifa, fill = "grey90", colour = "grey70", linewidth = 0.1) +
  geom_sf(data = filter(sal_price_map, fuelType == "U91"),
          aes(fill = median_price), colour = "grey70", linewidth = 0.1) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, na.value = "grey90") +
  coord_sf(xlim = c(melb_bbox["xmin"], melb_bbox["xmax"]),
           ylim = c(melb_bbox["ymin"], melb_bbox["ymax"])) +
  theme_bw() +
  labs(fill = "Median Price (c/L)")
```

```{r}
#| label: fig-dsl-map
#| fig-cap: "Median diesel price by suburb/locality within 50km of Melbourne CBD"

ggplot() +
  geom_sf(data = sal_seifa, fill = "grey90", colour = "grey70", linewidth = 0.1) +
  geom_sf(data = filter(sal_price_map, fuelType == "DSL"),
          aes(fill = median_price), colour = "grey70", linewidth = 0.1) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, na.value = "grey90") +
  coord_sf(xlim = c(melb_bbox["xmin"], melb_bbox["xmax"]),
           ylim = c(melb_bbox["ymin"], melb_bbox["ymax"])) +
  theme_bw() +
  labs(fill = "Median Price (c/L)")
```

```{r}
#| label: tbl-u91-cheapest
#| tbl-cap: "10 cheapest suburbs for U91 within 50km of Melbourne CBD"

sal_price_map_sa4 |>
  st_filter(melb_50km) |>
  st_drop_geometry() |>
  filter(fuelType == "U91") |>
  slice_min(median_price, n = 10) |>
  arrange(median_price) |>
  select(SAL_name, SA4_NAME21, median_price) |>
  rename(Suburb = SAL_name, `SA4 Region` = SA4_NAME21, `Median Price (c/L)` = median_price) |>
  knitr::kable(digits = 1)
```

```{r}
#| label: tbl-dsl-cheapest
#| tbl-cap: "10 cheapest suburbs for diesel within 50km of Melbourne CBD"

sal_price_map_sa4 |>
  st_filter(melb_50km) |>
  st_drop_geometry() |>
  filter(fuelType == "DSL") |>
  slice_min(median_price, n = 10) |>
  arrange(median_price) |>
  select(SAL_name, SA4_NAME21, median_price) |>
  rename(Suburb = SAL_name, `SA4 Region` = SA4_NAME21, `Median Price (c/L)` = median_price) |>
  knitr::kable(digits = 1)
```
The cheapest U91 suburbs are scattered across different directions from the CBD — north, south, and southeast, suggesting distance alone isn't the deciding factor. If geography were the main driver, we'd expect a tighter cluster. Diesel tells a different story: eight of the ten cheapest suburbs sit in Melbourne's west, hinting at something structural in that corridor.
So what do these suburbs have in common? One likely answer is competition — specifically, the presence of high-volume, price-aggressive operators. if location doesn't fully explain price variation, could socioeconomic factors be at play? Are disadvantaged communities quietly paying more at the pump?"

**Is Socioeconomic the reason?**


```{r}
#| label: fig-seifa-all-dotplot
#| fig-cap: "Median fuel price vs IRSD decile for all suburbs with data (1 = most disadvantaged)"

sal_price_map_sa4 |>
  st_drop_geometry() |>
  filter(!is.na(median_price), !is.na(IRSD_decile)) |>
  ggplot(aes(x = as.numeric(IRSD_decile), y = median_price)) +
  geom_point(alpha = 0.4, size = 1.5, colour = "steelblue") +
  scale_x_continuous(breaks = 1:10) +
  facet_wrap(~ fuelType, scales = "free_y") +
  theme_bw() +
  labs(
    x = "IRSD Decile (1 = Most Disadvantaged, 10 = Least Disadvantaged)",
    y = "Median Price (c/L)"
  )
```

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.
:::
